Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-04 text-backup.tex
Finalizing Discussion 85%

>>>>

    %%%%%%%DISCUSSION BACKUP REWRITE MAY "="%%%%%%%%

\section{Discussion}

This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is in turn influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
% - a few sentences on basic seasonal dynamics removed here
These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}). After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries

% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}). Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.

When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}). When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. \citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories. Here, we focus on the more consistently sampled fluorometric measurements. Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}. It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.

Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004. Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}. This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}. Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.

% check sizes of cells, then work this into a paragraph with another sentence!!
If we compare the two parts of the first cluster, the community reorganized, particularly as diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.

However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts. Instead, we observe a marked reduction following 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014. The same trend is evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. %

The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community. This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}. We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels. 
This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
We observe a significant difference in the Pilou index between the two clusters. Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}. This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community. The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$). As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictors of changes within the community are not in situ measurements, but rather the AMO and the MEI v.2 climate indices. These are followed in importance by nitrate concentration and SST (see Figure \ref{fig:GF}).

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2. This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1. Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.

Using microscopy cell counts as a basis for this analysis meant that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitation of identifying smaller species via light microscopy. Grouping cell counts by genera does create a loss of resolution, but is based on the assumption that species within a genus generally show similar ecophysiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, but also some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$). Within these groups, we see differing threshold responses in cumulative importance along the predictor ranges.

%AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. % COPIED TO INTRO
Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south. This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns. Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies. This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin. The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.

% Move the general point on ENSO to intro! Only Discussion-relevant part here..
%copied to Intro: The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression. In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag. We also tested a 12 month lag and found no significant improvement.
The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions. For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events. Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range. Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).

The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, SST and salinity, in that order. Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups. Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling. Similarly, for temperature, dinoflagellates and diatoms exhibit a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling. For the other predictors, we do not observe strong threshold responses.

Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco basin. This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances. One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month. Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community. However, we do not see such strong predictive power in the ERA5 climate reanalysis-derived mean wind speed, which also aggregates measurements and model output over an entire month. 
Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$), where both MEI v.2 and AMO scale along the second axis (NMDS 2). %TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.

% DEFINITELY NEED TO DISCUSS PROBLEMS OF CORRELATION WITH SUCH TIME SERIES DATA

% FROM FRAKER ET AL. 2022
%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships. As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."

- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}. % <- for this paper, check references to ENSO MEI v.2!!! This could be interesting!

Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}. In the Cariaco basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplaknton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}. Due to the lack of temporal resolution for higher trophic levels and the lack of zooplankton data for the first 6 years of the time series, we propose that mechanistic modeling studies could be well suited to investigate this aspect of the CARIACO time series further.

In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin. Despite multiannual trends in warming, reduced productivity and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions returned. We see these patterns clearly in the community data retrieved from microscopy, which shows the value of the arduous work of identifying and counting phytoplankton cells. The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles. We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.

% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters). Probably need more data, other approaches to untangle effects on diversity.
